\pkg{cleanR}: Maid for Cleaning Datasets in \proglang{R}

### 2016-12-29 latex/article_vol2.tex
Second bunch of commits

Though adhering to the template above is sufficient for using the freshly made \R{mySummaryFunction()} in \R{clean()}, we recommend furthermore adding it to the overview of all summary functions by converting it to a proper \code{summaryFunction} object. This is done by writing
\begin{Verbatim}
mySummaryFunction <- summaryFunction(mySummaryFunction,
	description = "[Some text describing what the summaryFunction does]",
	classes = c([the data types that this function is intended to be used for]))
\end{Verbatim}
which adds the new function to the output of an \code{allSummaryFunctions()} call. One comment should be devoted to the two attributes of a \R{summaryFunction}. If the \R{description} argument is left unspecified, the name of the function (in this case, \R{"mySummaryFunction"}) will be filled in. What happens if the \R{classes} argument is not specified depends on the type of \R{mySummaryFunction}. If \R{mySummaryFunction} is a \R{S3} generic function with associated methods, the call to \R{summaryFunction()} will automatically produce a vector of the names of the classes for which the function can be called. If \R{mySummaryFunction} is not an \R{S3} generic and \R{classes} is left unspecified, the attribute will simply be empty. Note that the helper function \R{allClasses()} might be useful for filling out the \R{classes} argument, as it simply lists all available classes in \pkg{cleanR}:
\begin{Schunk}
\begin{Sinput}
> allClasses()
\end{Sinput}
\begin{Soutput}
[1] "character" "Date"      "factor"   "integer"   "labelled"  
[6] "logical"  "numeric"  
\end{Soutput}
\end{Schunk}
\hl{Write something here, don't end paragraph with code. Also, maybe move the allClasses() stuff somewhere else, it doesn't really belong under this header. Not sure where to, though.}

\subsubsection{Writing a visualFunction}
\code{visualFunction}s are the functions that produce the figures of a \pkg{cleanR} output document. Writing a \R{visualFunction} is slightly more complicated than writing a \R{summaryFunction}. This follows from the fact that \R{visualFunctions} need to be able to output standalone code for plots in order for \code{clean()} to build standalone \R{rmarkdown} files. We recommend using the following structure:
\begin{Verbatim}
myVisualFunction <- function(v, vnam, doEval) {
  thisCall <- call("[the name of the function used to produce the plot]",
    v, [additional arguments to the plotting function])
  if (doEval) {
    return(eval(thisCall))
  } else return(deparse(thisCall))
}

myVisualFunction <- visualFunction(myVisualFunction,
	description = "[Some text describing the visualFunction]",
	classes = c([the data types that this function is intended to be used for]))
	)
\end{Verbatim}
In this function, \code{v} is the variable to be visualized, \code{vnam} is its name (which should generally be passed to \code{title} or \code{main} arguments in plotting functions) and \code{doEval} controls whether the output is a plot (if \code{TRUE}) or a character string of standalone code for producing a plot (if \code{FALSE}). Implementing the \code{doEval = TRUE} setting is not strictly necessary for a \R{visualFunction}'s use in \code{clean}, but it makes it easier to assess what visualization options are available, and obviously, it is crucial for interactive usage of \R{myVisualFunction()}. In either case, it should be noted that all the parameters listed above, \code{v}, \code{vnam} and \code{doEval}, are mandatory, so they must be left as is, even if they are not in use.

\subsubsection{Writing a checkFunction}
The last, but perhaps most important, \pkg{cleanR} function type is the \code{checkFunction}. These are the functions that flag issues in the data in the check step and control the overall flow of the data cleaning process in the precheck stage. A \code{checkFunction} can follow two overall structures, depending on the type of check. Either, it tries to identify problematic values in the variable (as e.g. \R{identifyMissing()} does) or it performs a check concerning the variable as a whole (e.g. the functions used for prechecks and the function \R{identifyNums()}). We present templates for both types of \R{checkFunction}s below separately, but it should be emphasized that formally, they belong to the same class.

First, a template for the full-variable check function type:
\begin{Schunk}
\begin{Sinput}
myFullVarCheckFunction <- function(v, ...) {
  [do your check]
  problem <- [is there a problem? TRUE/FALSE]
  message <- "[message describing the problem, if any]"
  checkResult(list(problem = problem,
    message = message,
    problemValues = NULL))
}

myFullVarCheckFunction <- checkFunction(myFullVarCheckFunction, 
  description = "[Some text describing the checkFunction]",
  classes = c([the data types that this function is intended to be used for])
  )
\end{Sinput}
\end{Schunk}
Again, as with \R{summaryFunction}s and \R{visualFunction}s, the change of function class by use of \R{checkFunction()} is not strictly necessary. Note however, that if \R{myFullVarCheckFunction} is to be used in the SVC step in \R{clean()}, the description attribute will be printed in the overview table in the \textit{Data cleaning summary}.

@@

	problemMessage <- "[The message that should be printed prior to listing
			problem values in the cleanR output, ending with a colon]"

@@

myProbValCheckFunction <- checkFunction(myProbValCheckFunction,
  description = "[Some text describing the checkFunction]",
  classes = c([the data types that this function is intended to be used for])
  )
\end{Verbatim}
In this template, the argument \R{maxDecimals} is not in use. This argument should be used to round off the \R{problemValues} passed to \R{messageGenerator()}, if they are numerical. This is done by substituting the \R{problemStatus} assignment above with the following code:
\begin{Schunk}
\begin{Sinput}
problemStatus <- list(problem = problem, 
  problemValues = round(problemValues, maxDecimals))
\end{Sinput}
\end{Schunk}
Another noteworthy component of the template is the usage of the helper function \R{messageGenerator()}, which aids consistent styling of all \code{checkFunction} messages. This function simply pastes together the \code{problemMessage} and the \code{problemValues}, with the latter being quoted and sorted alphabetically. If \R{nMax} is not \R{Inf}, only the first \R{nMax} problem values will be pasted onto the message, accompanied by a comment about how many problem values were left out (if any).  Note that printing quotes in rmarkdown requires an extensive amount of character escaping, so opting for \code{messageGenerator()} really is the easiest solution.

\subsection{A worked example}
\begin{figure}
\begin{center}
\includegraphics[scale=0.2]{mosaicPlotExample.pdf}
\end{center}
\label{fig:mosaicPlot}
\caption{\hl{something, note that code for producing this plot is available in ./latex/codeForArticle.R}}
\end{figure}

We will now build four new functions and show both how they can be used interactively and how they can be integrated with the \R{clean()} function. These four new functions are:
\begin{description}
\item[\R{isID}] A new \R{checkFunction} intended for use in the precheck-stage. This function checks whether a variable consists exclusively of long ($> 10$ characters/digits) entries that are all of equal length, as this might be personal identification codes that we do not wish to print out in the data summary.  
\item[\R{mosaicVisual}] A new \R{visualFunction} that produces so-called mosaic plots. This function will be used in the \textit{visualize} step of \R{clean()}.
\item[\R{countZeros}] A new \R{summaryFunction} that counts the number of occurrences of the value \R{0} in a variable. This function will be used in the \textit{summarize} step of \R{clean()}.
\item[\R{identifyColons}] A new \R{checkFunction} that flags variables in which values have colons that appear before and after alphanumerical characters. This is e.g. practical for identifying autogenerated interaction effects. This function will be used in the \textit{check} step of \R{clean()}. 
\end{description}
These functions are defined in turn below, and afterwards, an example of how they can be called from \R{clean()} is provided.

\subsubsection{isID - a new checkFunction without problem values}
First, let's define the \R{isID} function. As this function is not supposed to list problematic values in the variable, it falls within the category of \R{checkFunctions} represented by \R{myFullVarCheckFunction()} in the above. We do not particularly wish to use this function interactively, so we will stick to the minimal requirements of a \R{checkFunction} used in \R{check()} (see Table~\ref{table.functionTypes}). The function can then be defined by
\begin{Schunk}
\begin{Sinput}
isID <- function(v, nMax = NULL, ...) {
  out <- list(problem = FALSE, message = "")
  if (class(v) %in% setdiff(allClasses(), c("logical", "Date"))) {
    v <- as.character(v)
    lengths <- c(nchar(v))
    if (all(lengths > 10) & length(unique(lengths)) == 1) {
      out$problem <- TRUE
      out$message <- "Warning: This variable seems to contain ID codes."
    }
  }
  out
}
\end{Sinput}
\end{Schunk} 
\hl{Mention somewhere that  \R{mosiacplot()} is a base-R (\R{graphics}) function? Otherwise, it might not be completely clear, that we are passing a function name in the above...} This is essentially all we need to do in order to include this function as a precheck-function in \R{clean()}, so we will leave it as is and move on to the next function, namely \R{mosaicVisual}.

\subsubsection{mosiacVisual - a new visualFunction}
We will define this function such that it gets the full \pkg{cleanR} functionality. This can be done using the code 
\begin{Schunk}
\begin{Sinput}
 mosaicVisual <- function(v, vnam, doEval) {
   thisCall <- call("mosaicplot", table(v), main = vnam, xlab = "")
   if (doEval) {
    return(eval(thisCall))
   } else return(deparse(thisCall))
 }
\end{Sinput}
\end{Schunk}
This function can now be called directly or used in \R{clean()}. We will return to its usage in \R{clean()} below. Depending on the argument \R{doEval}, either a text string with code or a plot is produced. The plot resulting from the following call is found in Figure~\ref{fig:mosaicPlot}:
\begin{Schunk}
\begin{Sinput}
mosaicVisual(toyData$var1, "variable 1", doEval = TRUE)   $
\end{Sinput}
\end{Schunk}
\hl{remove \$.}
Even though \R{mosaicVisual}, as written above, follows the style of a \R{visualFunction}, it is not yet truly one and therefore, it will not appear in a \R{allVisualFunctions()} call. In order to get this functionality, we need to change its object class. This can be done by writing
\begin{Schunk}
\begin{Sinput}
 mosaicVisual <- visualFunction(mosaicVisual, 
 				description = "Mosaic plots using graphics",
                 classes = allClasses())
\end{Sinput}
\end{Schunk}
Here, we use the function \R{allClasses()} to quickly obtain a vector of all the seven variable classes addressed in \pkg{cleanR}. Note that if \R{mosaicVisual} were an S3 generic function, this argument could have been left as \R{NULL} and then the classes for which methods are available would be added automatically. \hl{I'm repeating myself here, but I think it is quite a neat feature, so maybe that's okay?}

@@

basicVisual    Histograms and barplots using                            
               graphics

standardVisual Histograms and barplots using                            
               ggplot2                                                  
------------------------------------------------------------------------
\end{Soutput}
\end{Schunk}
\hl{Redo this output table when makeXFunction issues are fixed.}

Now, we are done with the definition of \R{mosaicVisual} and we can turn to the next function in line, \R{countZeros}.

\subsubsection{countZeros - a new summaryFunction}
This \R{summaryFunction} in spe is defined in the following lines of code:
\begin{Schunk}
\begin{Sinput}
countZeros <- function(v, ...) {
	res <- length(which(v == 0))
	summaryResult(list(feature = "No. zeros", result = res, value = res))
}
\end{Sinput}
\end{Schunk}
Note that as this function computes an integer (the number of zeros), there is no difference between the entires \R{\$result} and \R{\$value}. If, on the other hand, the result had been a character string, extra formatting might be required in the \R{\$result} entry (such as escaping of quotation marks), and in this scenario, the two entries would then differ. As the result is returned as a \R{summaryResult} object, a printing method is automatically called when \R{countZeros} is used interactively:
\begin{Schunk}
\begin{Sinput}
> countZeros(c(rep(0, 5), 1:100))
\end{Sinput}
\begin{Soutput}
No. zeros: 5
\end{Soutput}
\end{Schunk}
As with \R{mosaicVisual()}, we change the class of this function in order to make it appear in \R{allSummaryFunctions()} calls. But now we wish to emphasize that the function is not intended to be called on all variable types, as zeros have different roles in \R{Date}s and in \R{logical} variables:
\begin{Schunk}
\begin{Sinput}
> countZeros <- summaryFunction(countZeros,
	description = "Count number of zeros",
	classes = c("character", "factor", "integer", 
	  "labelled", "numeric"))
\end{Sinput}
\end{Schunk}
\hl{more? don't end on code.}

\subsubsection{identifyColons - a new checkFunction with problem values}
The last function mentioned above is \R{identifyColons()}. We define it using the helper function \R{messageGenerator} to obtain a properly escaped message, and we use \R{checkResult} to make its output print neatly: 
\begin{Schunk}
\begin{Sinput}
identifyColons <- function(v, nMax = Inf, ... ) {
  v <- unique(na.omit(v))
  problemMessage <- "Note: The following values include colons:"
  problem <- FALSE
  problemValues <- NULL

  problemValues <- v[sapply(gregexpr("[[:xdigit:]]:[[:xdigit:]]", v),
                            function(x) all(x != -1))]

@@

  problemStatus <- list(problem = problem, 
                        problemValues = problemValues)
  outMessage <- messageGenerator(problemStatus, problemMessage, nMax)

@@

identifyColons <- checkFunction(identifyColons, 
    description = "Identify colons surrounded by alphanumeric characters",
    classes = c("character", "factor", "labelled"))

\end{Sinput}
\end{Schunk}
As with the previous two functions, we also change its class. Note, however, that for \R{checkFunction}s, the function description will appear in the document produced by \R{clean()} (in the \textit{Data cleaning summary} section), so now this is not only done for the sake of the \R{allCheckFunctions()} output.\\

\subsubsection{Calling the new SVC functions from clean()}
Now, we are ready to use these new functions in a \R{clean()} call. \hl{do this... what dataset should we use here? Include output in appendix, maybe.}. The extended \pkg{cleanR} output document should have the following modifications, relative to the standard \pkg{cleanR} output:
\begin{itemize}
\item We want to add the new pre-check function, \R{isID}, to the already existing pre-checks.
\item We wish to change the plot type for all variables to the new mosaic plot.
\item We want the new summary function, \R{countZeros}, to be added to the summaries performed on all variable types but \R{Date} and \R{logical}.
\item We want the new check function, \R{identifyColon}, to be added to the checks performed on \R{character}, \R{factor} and \R{labelled} variables.
\end{itemize}
These options are specified as follows:
\begin{Schunk}
\begin{Sinput}
> clean(dataSet, 
      preChecks = c("isKey", "isEmpty", "isID"),
      allVisuals = "mosaicVisual",
      characterSummaries = c(defaultCharacterSummaries(), "countZeros"),
      factorSummaries = c(defaultFactorSummaries(), "countZeros"),
      labelledSummaries = c(defaultLabelledSummaries(), "countZeros"),
      numericSummaries = c(defaultNumericSummaries(), "countZeros"),
      integerSummaries = c(defaultIntegerSummaries(), "countZeros"),
      characterChecks = c(defaultCharacterChecks(), "identifyColons"),
      factorChecks = c(defaultFactorChecks(), "identifyColons"),
      labelledCheck = c(defaultLabelledChecks(), "identifyColons"))
\end{Sinput}
\end{Schunk}
The outputted document is found in Appendix \hl{NUMTWO}. \hl{Comment. Make appendix. Remember to change dataSet to something else, when we decide what data to use here.}

\section{Something like examples}
Finally, we will present a few examples of how to make \pkg{cleanR} solve specific issues related to data cleaning. First, we discuss the challenges related to cleaning large datasets, particularly in terms of memory use and computation speed. Next, we show how \pkg{cleanR} can be used for problem-flagging. Lastly, we discuss how the \pkg{cleanR} output document can be included in other \R{rmarkdown} documents as a mean to produce clear and concise documentation of a dataset. \hl{I feel like there should be more topics here, but I'm all out of ideas...}

\subsection{Cleaning large datasets}
If the dataset becomes very large, the standard use of \R{clean()} outlined above might not be ideal. If there is a vast number of variables, production of the \R{rmarkdown} document might be quite slow, while an extensive amount of observations generally affects the rendering time of this document. In this section, we give a few practical examples of ways to deal with large data, while wishing to still produce (potentially very long) data cleaning overview documents. Note that the interactive tools of \pkg{cleanR} can be used as usual or sequentially in small subsets of the large dataset, if no such overview documents are needed.

\subsubsection{Attacking the figures}
Though figures give a nice overview of each variable, they are also quite heavy objects in terms of memory allocation. Therefore, it might be beneficial to not include figures in the \pkg{cleanR} outputs for very large datasets. This is controlled via the \R{mode} argument:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, mode = c("summarize", "check"))
\end{Sinput}
\end{Schunk}
If figures are indeed needed, a different approach is to choose the less memory heavy standard \R{R} figure style instead of the \R{ggplot2}\hl{reference?} figures that are the default option in \R{clean()}. This can be done using the \R{allVisuals} argument:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, allVisuals = "basicVisual")
\end{Sinput}
\end{Schunk}
Of course, even less heavy plots might be achieved by writing new \R{visualFunction}s, using the guidelines from section \ref{sec:functionTemplates}. For instance, a future extension of \pkg{cleanR} might be the inclusion of ASCII plots, as e.g. represented in the \R{R} package \R{txtplot}\hl{reference here? It's on CRAN.}. \\

\hl{I really feel like we should do some benchmarking here, maybe just on toyData, both in terms of speed and memory use. I would make the recommendations more trustworthy and serious.}

>>>>

Adhering to the template above is sufficient for using the freshly
made \R{mySummaryFunction()} in \R{clean()}, but we recommend
furthermore adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \R{summaryFunction} function with
the user-defined function as first argument, and additional arguments
\R{description} which adds explanatory flavour text as an attribute of
the function, and \R{classes} which determines which classes the
user-defined function should be applied to (also stored as an
attribute).
\begin{Verbatim}
mySummaryFunction <- summaryFunction(mySummaryFunction,
	description = "[Text describing what the summaryFunction does]",
	classes = c([vector of data types that the function is intended for]))
\end{Verbatim}
which adds the new function to the output of an
\code{allSummaryFunctions()} call. 
% One comment should be devoted to
% the two attributes of a \R{summaryFunction}. 
If the \R{description} argument is left unspecified, the name of the
function (which in this case is \R{"mySummaryFunction"}) will be filled in and
stored as an attribute. What happens if the \R{classes} argument is
not specified depends on the type of \R{mySummaryFunction}. If
\R{mySummaryFunction} is a \R{S3} generic function with associated
methods, the call to \R{summaryFunction()} will automatically produce
a vector of the names of the classes for which the function can be
called. If \R{mySummaryFunction} is not an \R{S3} generic and
\R{classes} is left unspecified, the attribute will simply be
empty. Note that the helper function \R{allClasses()} might be useful
for filling out the \R{classes} argument, as it simply lists all
available classes in \pkg{cleanR}:
\begin{Schunk}
\begin{Sinput}
> allClasses()
\end{Sinput}
\begin{Soutput}
[1] "character" "Date"      "factor"   "integer"   "labelled"  
[6] "logical"  "numeric"  
\end{Soutput}
\end{Schunk}
% \hl{Write something here, don't end paragraph with code. Also, maybe move the allClasses() stuff somewhere else, it doesn't really belong under this header. Not sure where to, though.}

If we run the \R{allSummaryFunctions()} function we can see that the
new function has been registered (output slightly appreviated):

\begin{Schunk}
\begin{Sinput}
> allSummaryFunctions()
\end{Sinput}
\begin{Soutput}
----------------------------------------------------------------------------
name                description                                      classes
------------------- --------------------------------------------------------
mySummaryFunction   [Text describing what the summaryFunction does]  numeric
.
.
.
\end{Soutput}
\end{Schunk}

\subsubsection{Writing a visualFunction}
\code{visualFunction}s are the functions that produce the figures in a
\pkg{cleanR} output document. Writing a \R{visualFunction} is slightly
more complicated than writing a \R{summaryFunction}. This follows from
the fact that \R{visualFunctions} need to be able to output standalone
code for plots in order for \code{clean()} to build standalone
\pkg{rmarkdown} files. We recommend using the following structure:
\begin{Verbatim}
myVisualFunction <- function(v, vnam, doEval) {
  thisCall <- call("[the name of the function used to produce the plot]",
    v, [additional arguments to the plotting function])
  if (doEval) {
    return(eval(thisCall))
  } else return(deparse(thisCall))
}
\end{Verbatim}

In this function, \code{v} is the variable to be visualized,
\code{vnam} is its name (which should generally be passed to
\code{title} or \code{main} arguments in plotting functions) and
\code{doEval} controls whether the output is a plot (if \code{TRUE})
or a character string of standalone code for producing a plot (if
\code{FALSE}). Implementing the \code{doEval = TRUE} setting is not
strictly necessary for a \R{visualFunction}'s use in \code{clean()}, but
it makes it easier to assess what visualization options are available,
and obviously, it is crucial for interactive usage of
\R{myVisualFunction()}. In either case, it should be noted that all
the parameters listed above, \code{v}, \code{vnam} and \code{doEval},
are mandatory, so they must be left as is, even if they are not in
use. See Table~\ref{table.functionTypes}.

Similarly to the summary function, we call \R{visualFunction()} to
register our newly created function. Like before, it accepts the
user-defined function as the first argument, and two additional
arguments description which adds explanatory flavour text as an
attribute of the function, and a vector which determines which classes
the user-defined function should be applied to (which is also stored
as an attribute).

\begin{Verbatim}
myVisualFunction <- visualFunction(myVisualFunction,
	description = "[Some text describing the visualFunction]",
	classes = c([data types that this function is intended for]))
	)
\end{Verbatim}

Again, registering the visualization function allows it to be used
without manually adding it to the list of visualizations.

\subsubsection{Writing a checkFunction}
The last, but perhaps most important, \pkg{cleanR} function type is
the \code{checkFunction}. These are the functions that flag issues in
the data in the check step and control the overall flow of the data
cleaning process in the precheck stage. A \code{checkFunction} follows
one of two overall structures, depending on the type of check. Either,
it tries to identify problematic values in the variable (as
e.g., \R{identifyMissing()} does) or it performs a check concerning the
variable as a whole (e.g. the functions used for prechecks and the
function \R{identifyNums()}). We present templates for both types of
\R{checkFunction}s below separately, but it should be emphasized that
formally, they belong to the same class.

First, a template for the full-variable check function type, where we
first define the function and subsequently register it as a check
function using \R{checkFunction()}:
\begin{Schunk}
\begin{Sinput}
myFullVarCheckFunction <- function(v, ...) {
  [do your check]
  problem <- [is there a problem? TRUE/FALSE]
  message <- "[message describing the problem, if any]"
  checkResult(list(problem = problem,
    message = message,
    problemValues = NULL))
}

myFullVarCheckFunction <- checkFunction(myFullVarCheckFunction, 
  description = "[Some text describing the checkFunction]",
  classes = c([the data types that this function is intended to be used for])
  )
\end{Sinput}
\end{Schunk}

Again, as with \R{summaryFunction}s and \R{visualFunction}s, the
change of function class by use of \R{checkFunction()} is not strictly
necessary. Note however, that if \R{myFullVarCheckFunction} is to be
used in the summarize/visualize/check steps in \R{clean()}, the
description attribute will be printed in the overview table in the
\textit{Data cleaning summary} part of the output document.

@@

	problemMessage <- "[Message that is printed prior to listing
			    problem values in the cleanR output, 
                            ending with a colon]"

@@

myProbValCheckFunction <- checkFunction(myProbValCheckFunction,
  description = "[Some text describing the checkFunction]",
  classes = c([the data types that this function is intended to be used for])
)
\end{Verbatim}
For this template, the argument \R{maxDecimals} is not in use. This
argument is used to round off the \R{problemValues} passed to
\R{messageGenerator()}, if they are numerical.  This is done by
substituting the \R{problemStatus} assignment above with the following
code: \hl{Denne foregående sætning forstår jeg ikke helt. Måske skal
  messagegenerator have en sætning eller to med på vejen? (Og det spiller lidt sammen med nedenstående)}
\begin{Schunk}
\begin{Sinput}
problemStatus <- list(problem = problem, 
   problemValues = round(problemValues, maxDecimals))
\end{Sinput}
\end{Schunk}

It is worth mentioning the 
%Another noteworthy component of the template is the usage of the
helper function \R{messageGenerator()}, which aids in consistent
styling of all \code{checkFunction} messages. This function simply
pastes together the \code{problemMessage} and the
\code{problemValues}, with the latter being quoted and sorted
alphabetically. If \R{nMax} is not \R{Inf}, only the first \R{nMax}
problem values will be pasted onto the message, accompanied by a
comment about how many problem values were left out (if any).  Note
that printing quotes in \pkg{rmarkdown} requires an extensive amount
of character escaping, so opting for \code{messageGenerator()} really
is the easiest solution.

\subsection{A worked example}

We will now build four new functions and show both how they can be
used interactively and how they can be integrated with the \R{clean()}
function. These four new functions are:
\begin{description}
\item[\R{isID}] A new \R{checkFunction} intended for use in the precheck-stage. This function checks whether a variable consists exclusively of long ($> 10$ characters/digits) entries that are all of equal length, as this might be personal identification codes that we do not wish to print out in the data summary.  
\item[\R{mosaicVisual}] A new \R{visualFunction} that produces mosaic plots. This function will be used in the \textit{visualize} step of \R{clean()}.
\item[\R{countZeros}] A new \R{summaryFunction} that counts the number of occurrences of the value \R{0} in a variable. This function will be used in the \textit{summarize} step of \R{clean()}.
\item[\R{identifyColons}] A new \R{checkFunction} that flags variables in which values have colons that appear before and after alphanumerical characters. This is practical for identifying autogenerated interaction effects. This function will be used in the \textit{check} step of \R{clean()}. 
\end{description}
These functions are defined in turn below, and afterwards, an example of how they can be called from \R{clean()} is provided.

\subsubsection{isID --- a new checkFunction without problem values}
First, let's define the \R{isID()} function. As this function is not
supposed to list problematic values in the variable, it falls within
the category of \R{checkFunctions} represented by
\R{myFullVarCheckFunction()} in the above. We do not particularly wish
to use this function interactively, so we will stick to the minimal
requirements of a \R{checkFunction} used in \R{check()} (see
Table~\ref{table.functionTypes}). The function can then be defined by
\begin{Schunk}
\begin{Sinput}
isID <- function(v, nMax = NULL, ...) {
  out <- list(problem = FALSE, message = "")
  # Should work for all but logical and Date variables
  if (class(v) %in% setdiff(allClasses(), c("logical", "Date"))) {
    v <- as.character(v)
    lengths <- c(nchar(v))
    # Check that all lengths are the same and greater than 10 characters
    if (all(lengths > 10) & length(unique(lengths)) == 1) {
      out$problem <- TRUE
      out$message <- "Warning: This variable seems to contain ID codes."
    }
  }
  out
}
\end{Sinput}
\end{Schunk}

This is essentially all we need to do in order to include this
function as a precheck-function in \R{clean()}, so we will leave it as
is and move on to the next function, namely \R{mosaicVisual}.

\subsubsection{mosiacVisual --- a new visualFunction}
\proglang{R} contains a function, \R{mosiacplot}, which produces
mosaic plots. We intend to use this existing high-level plotting
function as part of the new visualization function.  We will define
the new function such that it gets the full \pkg{cleanR}
functionality. This can be done using the following code.
\begin{Schunk}
\begin{Sinput}
 mosaicVisual <- function(v, vnam, doEval) {
   # Setup the call using the existing function mosaicplot
   thisCall <- call("mosaicplot", table(v), main = vnam, xlab = "")
   if (doEval) {                     # Return the graphics
    return(eval(thisCall))
   } else return(deparse(thisCall))  # Else return the code for the call
 }
\end{Sinput}
\end{Schunk}
This function can now be called directly or used in \R{clean()} and we
will show an example below. Depending on the \R{doEval} argument
\hl{skal doEval have en default value ovenfor?}, either a text string
with code or a plot is produced. The plot resulting from the following
call is found in Figure~\ref{fig:mosaicPlot}:
\begin{Schunk}
\begin{Sinput}
> mosaicVisual(toyData$var1, "variable 1", doEval = TRUE)   
\end{Sinput}
\end{Schunk}
% $

\begin{figure}
\begin{center}
\includegraphics[scale=0.2]{mosaicPlotExample.pdf}
\end{center}

\caption{Mosaic plot generated by the user-defined visualization
  function \texttt{mosaicVisual}.}
\label{fig:mosaicPlot}
\end{figure}

Even though \R{mosaicVisual()}, as written above, follows the style of a
\R{visualFunction}, it is not yet truly one and therefore, it will not
appear in a \R{allVisualFunctions()} call. In order to get this
functionality, we need to change its object class. This can be done by
writing
\begin{Schunk}
\begin{Sinput}
 mosaicVisual <- visualFunction(mosaicVisual, 
 				description = "Mosaic plots using graphics",
                                classes = allClasses())
\end{Sinput}
\end{Schunk}
Here, we use the function \R{allClasses()} to quickly obtain a vector
of all the seven variable classes addressed in \pkg{cleanR}. Note that
if \R{mosaicVisual} were an S3 generic function, this argument could
have been left as \R{NULL} and then the classes for which methods are
available would be added automatically. 
%\hl{I'm repeating myself here,
%  but I think it is quite a neat feature, so maybe that's okay?}

@@

basicVisual    Histograms and barplots using character, Date, factor,   
               graphics                      integer, labelled, logical,                           
                                             numeric                    
standardVisual Histograms and barplots using character, Date, factor,                              
               ggplot2                       integer, labelled, logical,               
                                             numeric                    
------------------------------------------------------------------------
\end{Soutput}
\end{Schunk}

Now that we are done with the definition of \R{mosaicVisual()} we can
turn to the next function in line, \R{countZeros}.

\subsubsection{countZeros --- a new summaryFunction}
This \R{summaryFunction} is defined in the following lines of code:
\begin{Schunk}
\begin{Sinput}
countZeros <- function(v, ...) {
	res <- length(which(v == 0))
	summaryResult(list(feature = "No. zeros", result = res, value = res))
}
\end{Sinput}
\end{Schunk}
As this function computes an integer (the number of zeros),
there is no difference between the entries \R{\$result} and
\R{\$value}. If, on the other hand, the result had been a character
string, extra formatting might be required in the \R{\$result} entry
(such as escaping of quotation marks), and in this scenario, the two
entries would have differed. As the result is returned as a
\R{summaryResult} object, a printing method is automatically called
when \R{countZeros} is used interactively:
\begin{Schunk}
\begin{Sinput}
> countZeros(c(rep(0, 5), 1:100))
\end{Sinput}
\begin{Soutput}
No. zeros: 5
\end{Soutput}
\end{Schunk}
As with \R{mosaicVisual()}, we change the class of this function in
order to make it appear in \R{allSummaryFunctions()} calls. But now we
wish to emphasize that the function is not intended to be called on
all variable types, as zeros have different roles in \R{Date}s and in
\R{logical} variables:
\begin{Schunk}
\begin{Sinput}
> countZeros <- summaryFunction(countZeros,
	description = "Count number of zeros",
	classes = c("character", "factor", "integer", 
                    "labelled", "numeric"))
\end{Sinput}
\end{Schunk}
% \hl{more? don't end on code.}

\subsubsection{identifyColons - a new checkFunction with problem values}
The last function mentioned above is \R{identifyColons()}. We define
it using the helper function \R{messageGenerator()} to obtain a properly
escaped message, and we use \R{checkResult} to make its output print
neatly:
\begin{Schunk}
\begin{Sinput}
identifyColons <- function(v, nMax = Inf, ... ) {
  v <- unique(na.omit(v))
  problemMessage <- "Note: The following values include colons:"
  problem <- FALSE
  problemValues <- NULL

  # Use regular expressions to identify colons between two words  
  problemValues <- v[sapply(gregexpr("[[:xdigit:]]:[[:xdigit:]]", v),
                            function(x) all(x != -1))]

@@

  problemStatus <- list(problem = problem, 
                        problemValues = problemValues)
  # Use the messagegenerator function to produce the output 
  # message from the problemStatus and problemMessage
  outMessage <- messageGenerator(problemStatus, problemMessage, nMax)

@@

identifyColons <- checkFunction(identifyColons, 
    description = "Identify colons surrounded by alphanumeric characters",
    classes = c("character", "factor", "labelled"))
\end{Sinput}
\end{Schunk}
As with the previous two functions, we also change its class. Note,
however, that for \R{checkFunction}s, the function description will
appear in the document produced by \R{clean()} (in the \textit{Data
  cleaning summary} section of the output), so now this is not only
done for the sake of the \R{allCheckFunctions()} output.

\subsubsection{Calling the new summarize/visualize/check functions from clean()}
Now, we are ready to use these new functions in a \R{clean()}
call. The extended \pkg{cleanR} output document should have the
following modifications, relative to the standard \pkg{cleanR} output:
\begin{itemize}
\item We want to add the new pre-check function, \R{isID}, to the already existing pre-checks.
\item We wish to change the plot type for all variables to the new mosaic plot.
\item We want the new summary function, \R{countZeros}, to be added to the summaries performed on all variable types but \R{Date} and \R{logical}.
\item We want the new check function, \R{identifyColon}, to be added to the checks performed on \R{character}, \R{factor} and \R{labelled} variables.
\end{itemize}
These options are specified as follows:
\begin{Schunk}
\begin{Sinput}
> data(testData) 
> clean(testData, 
      # Add the new pre-check function
      preChecks = c("isKey", "isEmpty", "isID"),
      # Change the visual
      allVisuals = "mosaicVisual",
      # Add the new checks for specific data types
      characterSummaries = c(defaultCharacterSummaries(), "countZeros"),
      factorSummaries = c(defaultFactorSummaries(), "countZeros"),
      labelledSummaries = c(defaultLabelledSummaries(), "countZeros"),
      numericSummaries = c(defaultNumericSummaries(), "countZeros"),
      integerSummaries = c(defaultIntegerSummaries(), "countZeros"),
      characterChecks = c(defaultCharacterChecks(), "identifyColons"),
      factorChecks = c(defaultFactorChecks(), "identifyColons"),
      labelledCheck = c(defaultLabelledChecks(), "identifyColons"))
\end{Sinput}
\end{Schunk}
The outputted document is found in Appendix \ref{sec:appendix2}.

\section{Using cleanR in specific situations}
Finally, we will present a few examples of how to make \pkg{cleanR}
solve specific issues related to data cleaning. First, we discuss the
challenges related to cleaning large datasets, particularly in terms
of memory use and computation speed. Next, we show how \pkg{cleanR}
can be used for problem-flagging. Lastly, we discuss how the
\pkg{cleanR} output document can be included in other \pkg{rmarkdown}
documents as a way to produce clear and concise documentation of a
dataset. %\hl{I feel like there should be more topics here, but I'm all
%  out of ideas...}

\subsection{Cleaning large datasets}
If the dataset becomes very large, the standard use of \R{clean()}
outlined above might not be ideal. If there is a vast number of
variables, creation of the \pkg{rmarkdown} document might be quite
slow, while a large number of observations generally will affect the
rendering time of the document. In this section, we give a few
practical examples of ways to deal with large data, while wishing to
still produce (potentially very long) data cleaning overview
documents. Note that the interactive tools of \pkg{cleanR} can be used
as usual or sequentially in small subsets of the large dataset, if no
such overview documents are needed.

\subsubsection{Handling the figures}
Though figures give a nice overview of each variable, they are also
quite heavy objects in terms of memory allocation. Therefore, it might
be beneficial to not include figures in the \pkg{cleanR} outputs for
very large datasets. This is controlled via the \R{mode} argument:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, mode = c("summarize", "check"))
\end{Sinput}
\end{Schunk}
If figures are indeed needed, a different approach is to choose the
less memory heavy standard \proglang{R} figure style instead of the
\pkg{ggplot2} figures that are the default option in \R{clean()}. This
can be done by setting the \R{allVisuals = "basicVisual"} argument:
\begin{Schunk}
\begin{Sinput}
> clean(toyData, allVisuals = "basicVisual")
\end{Sinput}
\end{Schunk}
Of course, even less heavy plots might be achieved by writing new
\R{visualFunction}s, using the guidelines from section
\ref{sec:functionTemplates}. For instance, a future extension of
\pkg{cleanR} might be the inclusion of ASCII plots, as
e.g. represented in the \proglang{R} package \pkg{txtplot}.

%\hl{I really feel like we should do some benchmarking here, maybe just on toyData, both in terms of speed and memory use. I would make the recommendations more trustworthy and serious.}
